A Study on Common Bug Fix Patterns in Rust

### 2022-11-02 discussion.tex
fixed discussion, results todos, added two G patterns, dropped one BC pattern

In this work, our pipeline successfully mined 19 cross project bug fix patterns. However, our pipeline fails to detect complex pattern due to space limitation and their innate complicated features. A complicated pattern is a pattern that includes many changes in many places and it makes it easier for two datapoints to be dissimilar.

Also, we expected to obtain certain patterns regarding some ertain Rust related features, such as unsafe Rust, but our results did not include those patterns. Our speculation is that, since Rust has a strong compiler check before creating the program executable, the programmers usually get many errors, and they only push their code once it is free of compile errors. That is why, we think there are some patterns that were hidden from our sight.

We believe our results can help researchers to create Rust tools for repair, fault localization, IDE linters, and etc. Also, these results can help program encoding methods to focus on more important aspects of a program and shrink down information more effectively.

>>>>

In this work, our pipeline successfully mined 20 cross project bug fix patterns. We introduced these patterns in two groups: 12 general patterns and 8 BC-related patterns. Clippy is able to detect four of the BC-related patterns. To the best of our knowledge, the rest of the patterns cannot be reported by the current linting tools. That is why, we believe our work introduces a new direction for IDE tools to help Rust developers.

In each group, we categorized the bug fix patterns based on the underlying program element. The patterns encompass a wide variety of program elements both in the general and BC-related patterns. This shows that our weighting scheme is acceptably unbiased. Also, the introduced patterns are all cross-project fix patterns. Each fix pattern has been seen at least in three different projects, indicating their commonality.

Furthermore, we expected to obtain certain patterns regarding some certain Rust related features, such as unsafe Rust, but our results did not include those patterns. Our speculation is that, since Rust has a strong compiler check before creating the program executable, the programmers usually get many errors, and they only push their code once it is free of compile errors. That is why, we think there are some patterns that were hidden from our sight.

We believe our results can help researchers to create Rust tools for repair, fault localization, IDE linters, and etc. Also, these results can help program embedding methods to focus on more important aspects of a program and shrink down information more effectively.

### 2022-11-02 results.tex
fixed discussion, results todos, added two G patterns, dropped one BC pattern

\todo[inline]{Can we justify the 50, both here and in the previous section?}

@@

\todo[inline]{no, you have to compute the intersection. when is our approach effective and when is it not? more detailed stats would probably be good too}

@@

We collected 60583 datapoints for general fix patterns using a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz) and the data collection took around 26 hours to compute both $D_g$ and $D_b$. We used $\epsilon=0.0001$ and $Z=5$ as our DBSCAN parameters and that resulted in 577 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 10 general patterns, which we introduce below.

@@

\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Attributes} & G.attr.struct & 172 & 16                                  \\
& G.attr.struct.field & 62 & 16                     \\\cline{1-4}
\multirow{2}{*}{Struct} & G.struct.field  & 179 & 17\\
& G.struct.field.pub & 23 & 12   
\\\cline{1-4}
Option & G.option & 22 & 12                    \\\cline{1-4}
\multirow{4}{*}{Types} & G.type.field  & 72 & 14 \\
& G.type.generic & 46 & 14\\ & G.type.enum.variant  & 39 & 12 \\
& G.type.tuple & 6 & 6                                         \\\cline{1-4}
\multirow{1}{*}{Traits} & G.trait.bound & 6 & 5 \\ 
\end{tabular}
\caption{\label{table:general}Bug Fix Pattern Categorization (General Patterns)}
\end{table}
\todo[inline]{There cannot be a table without reference to the table in the text. The reference to the table should also summarize what we've learned from the table. Also do we compare table 2 to what occurs in previous work?}

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, this modification also happens in three different ways. Using attributes for struct fields, apart from being able to store different meta-information for different struct fields, for some use cases, the programmers can control the action they want to apply on different fields.
\todo{this last sentence immediately prior to here has too many moving parts}

@@

\begin{lstlisting}[language=Rust, style=colouredRust]
pub struct Manifest {
  <@\textcolor{red}{--}@> pub features: Vec<String>,
  <@\textcolor{green}{++}@> pub features: HashSet<String>,
}

@@

\noindent\textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. For instance, the developer may be implementing a new feature or fixing the behaviour of the program. Also, it can happen for refactoring purposes or performance enhancement. \\
\todo{maybe discuss the specific example a bit?}

@@

\subsubsection{Traits}
\noindent\textit{\textbf{G.trait.bound} TraitBound Change}

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This change pattern relates to modifications in trait bounds. This can happen as part of bug fixes or for refactoring.

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 9 borrow-checker related fix patterns, which we introduce here:

\begin{table}[]
\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Clone}                                         & BC.clone.drop & 83 & 10 \\
& BC.clone.ref & 4 & 3   
\\\cline{1-4}
\multirow{3}{*}{Ref and Deref} & BC.ref.add & 17 & 7 \\
& BC.ref.drop & 145 & 14 \\
& BC.deref.to\_string  & 7 & 4                                            \\\cline{1-4}
\multirow{2}{*}{Mut} & BC.mut.add  & 11 & 6 \\
& BC.mut.drop  & 42 & 15                               \\\cline{1-4}
Vector & BC.vec.slice  & 8 & 6                                    \\\cline{1-4}
\multirow{1}{*}{Lifetime}      
& BC.lifetime.static  & 8 & 6
\\
\end{tabular}
\caption{\label{table:bc}Bug Fix Pattern Categorization (BC-Related Patterns)}
\end{table}

@@

\noindent\textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy\footnote{\url{https://github.com/rust-lang/rust-clippy}} is a linter tool for Rust that catches common mistakes. There are more than 500 lints included in Clippy. With regards to this pattern, Clippy detects whether a variable clone is necessary, i.e. could the developer have used the variable itself or not. This change does not modify the program behaviour, but it can boost the performance significantly if the program was repeatedly cloning large variables. \\

\todo[inline]{I think you should talk more about ``does not modify program behaviour''. For instance after you make this change the program may no longer compile. Is it true that, if it compiles, behaviour is preserved?}

@@

\noindent\textit{\textbf{BC.ref.drop} Removing Borrowing}

\begin{lstlisting}[language=Rust, style=colouredRust]
enums.iter().for_each(|e| {
    let mut enum_template = EnumTemplate::new();
    <@\textcolor{red}{--}@> enum_template.set_message_enum(&e);
    <@\textcolor{green}{++}@> enum_template.set_message_enum(e);
    // Other statements
});
\end{lstlisting}

\noindent\textbf{Description:} In this pattern, an unnecessary borrow is removed. Clippy can detect the needless borrows, and also can automatically remove them. This does not change the behaviour of the program, however, it does improve the performance, as the borrow checker need not to return the ownership after the execution of the function. \\
\todo[inline]{I don't see why this improves performance, it should be a 0 cost operation at runtime to return ownership or not.}

\noindent\textit{\textbf{BC.deref.to\_string()} Adding or removing a dereference operation before calling to\_string()}

@@

\todo[inline]{Can you clarify ``has been detected by Clippy''? Did you run Clippy? You didn't talk about that in your methodology.}
\noindent\textbf{Description:} This is a change that in all instances has been detected by Clippy. The change applies on an object prior to calling \verb+to_string()+ on that object. The change might add or remove a dereference, as shown above. When adding a dereference, the reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. It is faster and more performance efficient to dereference the object and then call \verb+to_string()+ on it. The deletion case indicates a redundant dereference, thus it is removed.\todo{I don't understand the adding change.}

>>>>

\todo[inline]{Can we justify the 50, both here and in the previous section? ``It was a random number not so large and not too small, how can I justify it?''}

@@

Although reporting most important element within the top three elements in 80\% of the cases is an accpectable outcome, we examined our pipeline for the other 20\%. The analysis shows that our pipeline fails in change patterns where the change involves many elements (e.g. adding a whole new \verb+impl+ block and implementing new functions in it). It is worth mentioning that these changes tend to be less cross-project as having many changed elements make it easier for datapoints to be dissimlar.

\todo[inline]{no, you have to compute the intersection. ``The numbers reported are the intersection'' when is our approach effective and when is it not? ``added above'' more detailed stats would probably be good too ``what kind of stats do you have in mind?''}

@@

We collected 60583 datapoints for general fix patterns using a server with a Xeon Gold 5120 processor (Q3-2017, 14-core 2.2 to 3.2GHz) and the data collection took around 26 hours to compute both $D_g$ and $D_b$. We used $\epsilon=0.0001$ and $Z=5$ as our DBSCAN parameters and that resulted in 577 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 10 general patterns, which we describe in Table~\ref{table:general}. As it is shown in the table, we have separated the patterns based underlying program element (Category column). We have given each pattern a specific ID (ID column) and determined the number of instances manifesting that fix pattern (\# Instances column), as well as the number of projects in which the pattern was seen (\# Project column). Now we provide detailed description for each of these patterns:

@@

\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Attributes} & G.attr.struct & 172 & 16                                  \\
& G.attr.struct.field & 62 & 16                     \\\cline{1-4}
\multirow{2}{*}{Struct} & G.struct.field  & 179 & 17\\
& G.struct.field.pub & 23 & 12   
\\\cline{1-4}
Option & G.option & 22 & 12                    \\\cline{1-4}
\multirow{4}{*}{Types} & G.type.field  & 72 & 14 \\
& G.type.generic & 46 & 14\\ & G.type.enum.variant  & 39 & 12 \\
& G.type.tuple & 6 & 6                                         \\\cline{1-4}
\multirow{1}{*}{Traits} & G.trait.bound & 6 & 5 \\\cline{1-4}
\multirow{2}{*}{Match} & G.match.pattern & 41 & 12 \\
& G.match.code & 8 & 5 \\
\end{tabular}
\caption{\label{table:general}Bug Fix Pattern Categorization (General Patterns)}
\end{table}
\todo[inline]{Also do we compare table 2 to what occurs in previous work? ``What previous work?''}

@@

\noindent\textbf{Description:} In this pattern, the attribute set of a struct field changes. As with \textit{\textbf{G.attr.struct}}, this modification also happens in three different ways. Using these attributes, the developers can control different actions they wish to apply on struct fields, in addition to providing meta-information.

@@

\begin{lstlisting}[language=Rust, style=colouredRust]
pub struct Manifest {
  <@\textcolor{red}{--}@> pub substitutions: HashMap<String, &'a str>,
  <@\textcolor{green}{++}@> pub substitutions: IndexMap<String, &'a str>,
}

@@

\noindent\textbf{Description:} The type of a struct field changes. This is a common change pattern in all statically-typed languages and can happen for various reasons. For instance, the developer may be implementing a new feature or fixing the behaviour of the program. Also, it can happen for refactoring purposes or performance enhancement.

For instance, in the Rust project starship/starship (a cross-shell prompt), in order to preserve the insertion order of \verb+substitutions+ field, the developer changes its type from \verb+HashMap+ to \verb+IndexMap+\footnote{\url{https://github.com/starship/starship/commit/4de9e43cff46c834bd340d24a02fc95d85310a33}}.
\\

@@

\subsubsection{Traits}
\noindent\textit{\textbf{G.trait.bound} TraitBound change}

@@

\noindent\textbf{Description:} The trait bounds are the functionalities that we require from our parametric types. This change pattern relates to modifications in trait bounds. This can happen as part of bug fixes or for refactoring. \\

\subsubsection{Match}

In Rust, match control flow construct is used for comparing a value against multiple patterns, and executing the code associated with the first pattern that matches with our value. A pattern and its associated code are called a match arm.\\

\noindent\textit{\textbf{G.match.pattern} Change in a match arm's pattern}

\begin{lstlisting}[language=Rust, style=colouredRust]
<@\textcolor{red}{--}@> Value::Primitive(p) => {
<@\textcolor{green}{++}@> UntaggedValue::Primitive(p) => {
    let _ = builder.add_empty_child(p.format(None));
}
\end{lstlisting}

\noindent\textbf{Description:} In this pattern, a change occurs in an arm's pattern. Like the code snippet provided above, it can be a change in pattern's type. It also can be matching to a different enum variant or string. The fix pattern can happen both for refactoring or debugging purposes. \\

\noindent\textit{\textbf{G.match.code} Change in a match arm's code}

\begin{lstlisting}[language=Rust, style=colouredRust]
Operation::Insert(insert) => {
    <@\textcolor{red}{--}@> inverted.delete(insert.count_of_utf16_code_units());
    <@\textcolor{green}{++}@> inverted.delete(insert.utf16_size());
}
\end{lstlisting}

\noindent\textbf{Description:} The block of code associated with a pattern in match arm can contain multiple statements. In this fix pattern, a change happens in at least one of these statements. Similar to the previous fix pattern, this also can be due to refactoring or debugging purposes.

@@

We collected 27143 datapoints for borrow-checker related fix patterns. We used $\epsilon=0.001$ and $Z=5$ as our DBSCAN parameters and that resulted in 102 clusters. After manual analysis, as described in~\ref{sec:manual_analysis_cluster_selection}, we ended up with 9 borrow-checker related fix patterns, which we, similar to the previous table, describe in Table~\ref{table:bc}. Here we provide detailed description for each of these patterns:

\begin{table}[]
\begin{tabular}{l|l|r|r}
\textbf{Catgegory} & \textbf{ID} & \textbf{\# Instances} & \textbf{\# Projects} \\
\hline
\multirow{2}{*}{Clone}                                         & BC.clone.drop & 83 & 10 \\
& BC.clone.ref & 4 & 3   
\\\cline{1-4}
\multirow{2}{*}{Ref and Deref} & BC.ref.add & 17 & 7 \\
& BC.deref.to\_string.add  & 4 & 3                                            \\\cline{1-4}
\multirow{2}{*}{Mut} & BC.mut.add  & 11 & 6 \\
& BC.mut.drop  & 42 & 15                               \\\cline{1-4}
Vector & BC.vec.slice  & 8 & 6                                    \\\cline{1-4}
\multirow{1}{*}{Lifetime}      
& BC.lifetime.static  & 8 & 6
\\
\end{tabular}
\caption{\label{table:bc}Bug Fix Pattern Categorization (BC-Related Patterns)}
\end{table}

@@

\noindent\textbf{Description:} This pattern is referred to as redundant clone in Clippy Lints. Clippy\footnote{\url{https://github.com/rust-lang/rust-clippy}} is a linter tool for Rust that catches common mistakes. There are more than 500 lints included in Clippy. With regards to this pattern, Clippy detects whether a variable clone is necessary, i.e. could the developer have used the variable itself or not. If the variable clone is indeed redundant, then removing it does not modify the program behaviour, however it can boost the performance significantly if the program was repeatedly cloning large variables. \\

\todo[inline]{I think you should talk more about ``does not modify program behaviour''. For instance after you make this change the program may no longer compile. Is it true that, if it compiles, behaviour is preserved? ``Yes the assumption is that if the presence of clone() is unnecessary, then removing it would not change the program behaviour''}

@@

\noindent\textit{\textbf{BC.deref.to\_string().add} Adding a dereference operation before calling to\_string()}

@@

\noindent\textbf{Description:} This is a change that in all instances has been detected by Clippy (commit messages indicated the use of Clippy for this fix). The change applies on an object prior to calling \verb+to_string()+ on that object. The change might add or remove a dereference, as shown above. When adding a dereference, the reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. Adding the dereference makes the compiler use the specialized implementation of \verb+to_string()+ and not go through slower string formatting methods.
